Drug sensitivity prediction with normal inverse Gaussian shrinkage informed by external data

### 2020-02-06 docs/supplement1.Rnw
commit before analysis_gdsc.R
added EMSE split for large and small betas to simulations_gdsc.R
added mcmc functions

% settings
\graphicspath{{../figs/}}
\pgfplotsset{compat=1.16}
\setlength{\evensidemargin}{.5cm} \setlength{\oddsidemargin}{.5cm}
\setlength{\textwidth}{15cm}
\title{Supplementary material to: Drug sensitivity prediction with the normal 
inverse Gaussian prior informed by external data}
\date{\today}
\author{Magnus M. M\"unch$^{1,2}$\footnote{Correspondence to: 
\href{mailto:m.munch@amsterdamumc.nl}{m.munch@amsterdamumc.nl}}, Mark A. van de 
Wiel$^{1,3}$, Sylvia Richardson$^{3}$, \\ and Gwena{\"e}l G. R. Leday$^{3}$}

@@

	\section{Normal inverse Gaussian prior moments}\label{sec:prior}
	Let $\beta \sim \mathcal{N}(0,\sigma^2 \tau^2 \gamma^2)$, where 
	$\sigma^2$ and $\tau^2$ are fixed and $\gamma^2 \sim 
	\mathcal{IG}(\phi,\lambda)$, then
	\begin{align*}
	  \E(\beta | \sigma^2, \tau^2) & = 0, \\
	  \V(\beta | \sigma^2, \tau^2) & = \sigma^2 \tau^2 \phi, \\
	  \mathcal{K}(\beta | \sigma^2, \tau^2) & = 
	  \frac{3\phi}{\lambda} + 3,
	\end{align*}
	where $\mathcal{K}(\cdot)$ denotes the kurtosis.

@@

  \section{Gibbs sampler}\label{sec:gibbssampler}
  MCMC samples from the posterior corresponding to model 
  (\ref{md-eq:prior}) may be generated for each equation independently. 
  MCMC samples may be generated by iteratively sampling the following 
  conditional distributions:
  \begin{subequations}
    \begin{align}
      \bm{\beta}_d | \bgamma_d^2, \sigma_d^2, \tau_d^2, \y_d & \sim 
      \mathcal{N}_p (\bm{\mu}_d, \bm{\Sigma}_d), \label{eq:betasample} \\
      \text{with } & \bm{\Sigma}_d = \sigma_d^2 [\X \tr \X + 
      \tau_d^{-2} \diag (\gamma_{jd}^{-2})]^{-1}, \label{eq:sigmasample} \\
      \text{and } & \bm{\mu}_d = [\X \tr \X + 
      \tau_d^{-2} \diag (\gamma_{jd}^{-2})]^{-1} \X \tr 
      \y_d \label{eq:musample}, \\
      \gamma_{jd}^2 | \beta_{jd}, \sigma_d^2, \tau_d^2, \y &
      \sim \mathcal{GIG}
      \(-1, \lambda_{\text{feat}}/ \phi_{jd}^{2}, \delta_{jd}\), \nonumber \\
      \text{with } & \delta_{jd} = \sigma_d^{-2} \tau_d^{-2} \beta_{jd}^2 + 
      \lambda_{\text{feat}}, \nonumber \\
      \tau_d^2 | \bm{\beta}_d, \sigma_d^2, \bgamma_{d}^2, \y & \sim
      \mathcal{GIG}
      \(-\frac{p+1}{2}, \lambda_{\text{drug}}/ \chi_{d}^{2}, \eta_d\), 
      \nonumber \\
      \text{with } & \eta_d = \sigma_d^{-2} \sum_{j=1}^p \gamma_{jd}^{-2}
      \beta_{jd}^2 + \lambda_{\text{drug}}, \nonumber \\
      \sigma_d^2 | \bm{\beta}_d, \bgamma_d^2, \tau_d^2, \y & \sim \Gamma^{-1} 
      \(\frac{n + p + 1}{2}, \zeta_d\), \nonumber \\
      \text{with } & \zeta_d = \frac{1}{2} \bigg[\mathbf{y} \tr \mathbf{y} -
      2 \mathbf{y} \tr \X \bm{\mu} + \trace ( \X \tr \X \bm{\Sigma}) + 
      (\bm{\mu}) \tr \X \tr \X \bm{\mu} + 
      \tau^{-2}_d \trace (\diag(\gamma_{jd}^{-2}) \bm{\Sigma}) \nonumber \\
      & \,\,\,\,\,\,\,\,\,\, + \tau_d^{-2} (\bm{\mu}) \tr 
      \diag(\gamma_{jd}^{-2}) \bm{\mu}\bigg]. \nonumber
    \end{align}
  \end{subequations}
  In high dimensional space, the $p \times p$ matrix inversions in 
  (\ref{eq:sigmasample}) and (\ref{eq:musample}) are   
  significant computational bottlenecks. \cite{bhattacharya_fast_2016} describe 
  a method to sample from (\ref{eq:betasample}) without explicit calculation
  of this inverse, thereby offering a siginifcant speed-up compared to naive
  sampling.

>>>>

% settings
\graphicspath{{../figs/}}
\pgfplotsset{compat=1.16}
\setlength{\evensidemargin}{.5cm} \setlength{\oddsidemargin}{.5cm}
\setlength{\textwidth}{15cm}
\title{Drug sensitivity prediction with normal inverse Gaussian shrinkage 
informed by external data}
\date{\today}
\author{Magnus M. M\"unch$^{1,2}$\footnote{Correspondence to: 
\href{mailto:m.munch@amsterdamumc.nl}{m.munch@amsterdamumc.nl}}, Mark A. van de 
Wiel$^{1,3}$, Sylvia Richardson$^{3}$, \\ and Gwena{\"e}l G. R. Leday$^{3}$}

@@

	\section{Normal inverse Gaussian prior moments}\label{sec:prior}
	Let $\beta \sim \mathcal{N}(0,\sigma^2 \tau^2 \gamma^2)$, where 
	$\sigma^2$ and $\tau^2$ are fixed and $\gamma^2 \sim 
	\mathcal{IG}(\phi,\lambda_{\text{feat}})$, then
	\begin{align*}
	  \E(\beta | \sigma^2, \tau^2) & = 0, \\
	  \V(\beta | \sigma^2, \tau^2) & = \sigma^2 \tau^2 \phi, \\
	  \mathcal{K}(\beta | \sigma^2, \tau^2) & = 
	  \frac{3\phi}{\lambda_{\text{feat}}} + 3,
	\end{align*}
	where $\mathcal{K}(\cdot)$ denotes the kurtosis. Now letting $\tau^2
	\sim \mathcal{IG}(\chi,\lambda_{\text{drug}})$ and integrating it out we have:
	\begin{align*}
	  \E(\beta | \sigma^2) & = 0, \\
	  \V(\beta | \sigma^2) & = \sigma^2 \chi \phi, \\
	  \mathcal{K}(\beta | \sigma^2) & = 
	  3\(\frac{\phi}{\lambda_{\text{feat}}} \frac{\chi}{\lambda_{\text{drug}}} +
	  \frac{\phi}{\lambda_{\text{feat}}} + \frac{\chi}{\lambda_{\text{drug}}}\) + 
	  3.
	\end{align*}

@@

  \section{Gibbs sampler}\label{sec:gibbssampler}
  MCMC samples from the posterior corresponding to model 
  (\ref{md-eq:prior}) may be generated for each equation independently.
  MCMC samples may be generated by iteratively sampling the following 
  conditional distributions:
  \begin{subequations}\label{eq:gibbssampler}
    \begin{align}
      \bm{\beta}_d | \bgamma_d^2, \sigma_d^2, \tau_d^2, \y_d & \sim 
      \mathcal{N}_p (\bm{\mu}_d, \bm{\Sigma}_d), \label{eq:betasample} \\
      \text{with } & \bm{\Sigma}_d = \sigma_d^2 [\X \tr \X + 
      \tau_d^{-2} \diag (\gamma_{jd}^{-2})]^{-1}, \label{eq:sigmasample} \\
      \text{and } & \bm{\mu}_d = [\X \tr \X + 
      \tau_d^{-2} \diag (\gamma_{jd}^{-2})]^{-1} \X \tr 
      \y_d \label{eq:musample}, \\
      \gamma_{jd}^2 | \beta_{jd}, \sigma_d^2, \tau_d^2, \y &
      \sim \mathcal{GIG}
      \(-1, \lambda_{\text{feat}}/ \phi_{jd}^{2}, \delta_{jd}\), \nonumber \\
      \text{with } & \delta_{jd} = \sigma_d^{-2} \tau_d^{-2} \beta_{jd}^2 + 
      \lambda_{\text{feat}}, \nonumber \\
      \tau_d^2 | \bm{\beta}_d, \sigma_d^2, \bgamma_{d}^2, \y & \sim
      \mathcal{GIG}
      \(-\frac{p+1}{2}, \lambda_{\text{drug}}/ \chi_{d}^{2}, \eta_d\), 
      \nonumber \\
      \text{with } & \eta_d = \sigma_d^{-2} \sum_{j=1}^p \gamma_{jd}^{-2}
      \beta_{jd}^2 + \lambda_{\text{drug}}, \nonumber \\
      \sigma_d^2 | \bm{\beta}_d, \bgamma_d^2, \tau_d^2, \y & \sim \Gamma^{-1} 
      \(\frac{n + p + 1}{2}, \zeta_d\), \nonumber \\
      \text{with } & \zeta_d = \frac{1}{2} \[\mathbf{y}_d \tr \mathbf{y}_d -
      2 \mathbf{y}_d \tr \X \bm{\beta}_d + \bm{\beta}_d \X \tr \X \bm{\beta}_d + 
      \tau^{-2}_d \bm{\beta}_d \tr \diag(\gamma_{jd}^{-2}) \bm{\beta}_d) \]. 
      \nonumber
    \end{align}
  \end{subequations}
  In high dimensional space, the $p \times p$ matrix inversions in 
  (\ref{eq:sigmasample}) and (\ref{eq:musample}) are   
  significant computational bottlenecks. \cite{bhattacharya_fast_2016} describe 
  a method to sample from (\ref{eq:betasample}) without explicit calculation
  of this inverse, thereby offering a siginifcant speed-up compared to naive
  sampling.

  Sampling from (\ref{eq:gibbssampler}) either requires specifying 
  hyperparameters
  $\bm{\alpha}_{\text{feat}}$, $\bm{\alpha}_{\text{drug}}$, 
  $\lambda_{\text{feat}}$, and $\lambda_{\text{drug}}$, or to endow them with
  and extra layer of hyperpriors. In general, specifying them requires 
  rarely available, detailed subject knowledge, so we resort to endowing them
  with vague (improper) hyperpriors:
  \begin{align*}
    \bm{\alpha}_{\text{feat}} & \sim c_1 \\
    \bm{\alpha}_{\text{drug}} & \sim c_2, \\
    \lambda_{\text{feat}} &\sim 1/\lambda_{\text{feat}}, \\
    \lambda_{\text{drug}} &\sim 1/\lambda_{\text{drug}},
  \end{align*}
  where $c_1$ and $c_2$ are constants. The corresponding full conditionals are
  given by:
  \begin{align*}
    \bm{\alpha}_{\text{feat}} | \lambda_{\text{feat}}, \bm{\Gamma} & \sim 
    \mathcal{N} \([\mathbf{C} \tr \diag(\gamma_{jd}^2) \mathbf{C}]^{-1}
    \mathbf{C} \tr \mathbf{1}_{pD \times 1}, \lambda_{\text{feat}}^{-1} 
    [\mathbf{C} \tr \diag(\gamma_{jd}^2) \mathbf{C}]^{-1} \) \\
    \bm{\alpha}_{\text{drug}} | \lambda_{\text{drug}}, \bm{\tau} & \sim 
    \mathcal{N} \([\mathbf{Z} \tr \diag(\tau_{d}^2) \mathbf{Z}]^{-1}
    \mathbf{Z} \tr \mathbf{1}_{D \times 1}, \lambda_{\text{drug}}^{-1} 
    [\mathbf{Z} \tr \diag(\tau_{d}^2) \mathbf{Z}]^{-1} \), \\
    \lambda_{\text{feat}} | \bm{\alpha}_{\text{feat}}, \bm{\Gamma} & 
    \sim \Gamma \(\frac{pD}{2}, \frac{1}{2} 
    \[\mathbf{C} \bm{\alpha}_{\text{feat}} - \diag(\gamma_{jd}^{-2}) 
    \mathbf{1}_{pD \times 1}\] \tr \diag(\gamma_{jd}^2) 
    \[\mathbf{C} \bm{\alpha}_{\text{feat}} - 
    \diag(\gamma_{jd}^{-2}) \mathbf{1}_{pD \times 1}\] \), \\
    \lambda_{\text{drug}} | \bm{\alpha}_{\text{drug}}, \bm{\tau} & 
    \sim \Gamma \(\frac{D}{2}, \frac{1}{2} 
    \[\mathbf{Z} \bm{\alpha}_{\text{drug}} - \diag(\tau_{d}^{-2}) 
    \mathbf{1}_{D \times 1}\] \tr \diag(\tau_{d}^2) 
    \[\mathbf{Z} \bm{\alpha}_{\text{drug}} - 
    \diag(\tau_{d}^{-2}) \mathbf{1}_{D \times 1}\] \).
  \end{align*}
